Future Cosmic Microwave Background delensing with galaxy surveys.

### 2017-06-29 Notes/galaxy_delensing.tex
fisher text and references

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{natbib}
\newcommand{\bx}{{\bf x}}
\newcommand{\bk}{{\bf k}}
\newcommand{\bq}{{\bf q}}
\newcommand{\bP}{{\bf\Psi}}
\newcommand{\bs}{{\bf s}}
\newcommand{\cs}{{\cal S}}
\newcommand{\by}{{\bf y}}
\newcommand{\deltar}{\delta_{\rm recon}}%\def\lcdm{$\Lambda$CDM}
\def\arcsec{$^{\prime\prime}$}
\def\nl{N_\ell}
\def\sl{S_\ell}
\def\ublu{\bl'}
\newcommand{\conv}[2]{\left(\frac{#1}{#2}\right)}
\def\arctanh{\mathop{\rm arctanh}\nolimits}
%\renewcommand{\eqref}[1] {equation $($\ref{#1}$)$}
\newcommand{\comment}[1]{{}}
\def\beq{\begin{equation}}
\def\eeq{\end{equation}}
\def\beqn{\begin{eqnarray}}
\def\eeqn{\end{eqnarray}}
\def\a{\alpha}
\def\cl{C_{l}}
\def\h{\mathrm{h}}
\def\d{\rmn{d}}
\def\pa{\partial}
\def\deldel#1#2{\frac{\pa{#1}}{\pa{#2}}}
\def\ba{\bm{\alpha}}
\def\fracj#1#2{{\textstyle{#1\over#2}}}
\def\bxi{\bm{\xi}}
\def\half{\frac{1}{2}}
\def\ti{\widetilde}
\def\O{\Omega}
\def\OL{\Omega_\Lambda}
\def\Om{\ensuremath{\Omega_{\mathrm{m}}}}
\def\Ob{\ensuremath{\Omega_{\mathrm{b}}}}
\def\Oc{\ensuremath{\Omega_{\mathrm{CDM}}}}
\def\msol{\ensuremath{M_\odot}}
\def\l{\left}
\def\r{\right}
\def\o{\omega}
\def\gcm{\textrm{g cm$^{-3}$}}
\def\2gcm{\textrm{g cm$^{-2}$}}
\def\Scr{\Sigma_{\mathrm{crit}}}
\def\rcr{\rho_{\mathrm{crit}}}
\def\phidot{\ensuremath{\dot\phi_{\bl,\bl'}}}
\def\ddelta{\ensuremath{\dot\delta}}
\def\modu#1{\l |{#1}\r |}
\def\av#1{\l \langle{#1}\r \rangle}
\def\hmpc{\:{h}^{-1}\mathrm{Mpc}}
\def\th{\Theta}
\def\tth{\tilde\Theta}
\def\sg{\sigma}
\def\Sig{\Sigma}
\def\cf{{\cal F}}
\def\k{\kappa}
\def\P{{P}}
\def\pnl{{P}_{{\!\textrm{\tiny NL}}}}
\def\dnl{\Delta_{\text{\scriptsize NL}}}
\def\kmin{\k_{\mathrm{min}}}
\def\kmax{\k_{\mathrm{max}}}
\def\hires{the \emph{highRes} experiment}
\def\lores{the \emph{lowRes} experiment}
\def\ktot{\k_{\mathrm{tot}}}
\def\hunit{\ensuremath{\mathrm{km}{\mathrm{s}^{-1}} \mathrm{Mpc}^{-1}}}
\def\H0{\ensuremath{\mathrm{H}_0}}
\def\nn{\nonumber}
\def\lin{\mathrm{lin}}
\def\ISW{\mathrm{ISW}}
\def\bl{\bmm{l}}
\def\bL{\bmm{L}}
\def\fsky{f_{\mathrm{sky}}}
\newcommand{\E}[1]{\times 10^{#1}}
\newcommand{\bmm}[1]{{\mathbf{#1}}}
\newcommand{\bsection}[1]{\section{\uppercase{#1}}}

@@

For future experiments, we need to increase the delensing efficiency by almost a factor of 3 to fully exploit the expected instrumental capabilities \cite{abazajian:2016}.
In this paper, we propose and study a possible way: using future galaxy surveys as tracers of the lensing potential in addition to other probes such as the SKA radio continuum survey, cosmic infrared background and internal CMB reconstruction. 
Furthermore, we point out how using redshift information through tomographic binning can improve the delensing efficiency of galaxy survey.
These will translate into a better reconstruction of the B-modes in the measured patch, and, as a consequence will improve the constraints on inflation trough delensing. We model several actual and future surveys, and after computing the residual B-modes, we forecast the resulting statistical uncertainties on the amplitude and the shape of the inflationary tensor perturbations for current CMB experiments as well as the next generation (S3) and the planned fourth generation (S4).

@@

The CMB lensing potential is \textit{the} field that we need to reconstruct to reverse the effect of large scale structure and delens the CMB.
However, the lensing potential can also be reconstructed using the CMB itself. In that case, we can treat it as a noisy tracer of the true field. Both the CMB lensing field and its noisy reconstructed counterpart have the same kernel $W^{\kappa}(z)$.
However, when computing the power spectrum of the latter, we need to add a noise component. Given the instrumental noise level and the beam, we can calculate the reconstruction noise $N_l^{\kappa \kappa}$, and so
\beq
\cl^{\kappa_\mathrm{rec} \kappa_\mathrm{rec}} = \cl^{\kappa \kappa } +N^{\kappa \kappa}_l
\eeq
In this work, the level of noise is computed assuming an iterative approach to the CMB lensing reconstruction as described in \cite{smith:2012,hirata:2003}.

@@

From this we get the power spectrum of the lensing component of the B-modes:
\be
C_l^{BB,\mathrm{lens}}  = \int \frac{d^2 \bl'}{(2 \pi)^2} W^2 (\bl,\bl')C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}.
\ee

The B-mode power spectrum measured on the sky is composed of a possible primordial component $C_l^{BB,r}$ together with the lensing $C_l^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_l^{BB}$ (defined in \refeq{noise}):
\be
C_l^{BB,\mathrm{full}} = C_l^{BB,r} + C_l^{BB,\mathrm{lens}} + N_l^{BB}.
\ee
The lensing component is a significant source of B-modes that, at large scales, correspond to a white noise source of roughly $5 \mu K$-arcmin. This means that it is not only bigger than the allowed inflationary component at scales smaller than several degrees, but it is also comparable with current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, we build a template \refeq{blens} of the lensing B-modes in the observed patch given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we can estimate $\phi$ using "tracers" of the dark matter distribution that sources the potential.

@@

The optimal weights $f(\bl,\bl')$ such that the residual lensing B mode power is minimized are:
\be
\label{eqn:fweight}
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }.
\ee
Here $C^{\kappa I}$ and $C^{II}$ are the cross correlation spectrum of the tracer $I$ with the lensing convergence $\kappa$ and its autospectrum; they are described for each LSS field in \refsec{model}. The power spectrum of the E-modes noise $N^{EE}$ is the same as the B-modes one in \refeq{noise}.

@@

With this choice of $f(\bl,\bl')$ we finally have that the residual power is:
\beqn
\label{eqn:Bres}
C_{{l}}^{BB,\mathrm{res}} &=&   \int \frac{d^2 \bl'}{(2 \pi)^2}  W^2 (\bl,\bl')
 C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}  \\
&\times&  \left[1 - \left(\frac{C^{EE}_{{l'}} }{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right) \rho^2_{|\bl-\bl'|} \right] \nonumber
\eeqn
with
\be
\rho_l= \frac{\cl^{\kappa I}}{\sqrt{\cl^{\kappa \kappa} \cl^{I I}}}.
\ee
The bigger the $\rho_l$ is for an LSS field, the more it is correlated with the lensing potential acting on the CMB photons. An higher correlation allows for a better reconstruction of $\phi$ and, as a consequence, of $B^{\rm{lens}}$.

@@

We start by assuming that we have n different tracers of the gravitational potentials $I_{i}$ with $i\in \{1,..,n\}$. We can optimally combine them to estimate $\phi$ or, in other word, maximizing the correlation factor $\rho$ with:
\beqn
I         &=&    \sum_{i}c^{i}I^{i} \nonumber \\
c_{i}     &=& (C_{II}^{-1})_{ij}C^{\kappa I^{j}}
\label{eqn:combined}
\eeqn
where $C_{II}$ is the covariance matrix of the LSS tracers.
The residual B-mode power can be derived from \refeq{Bres} using an ``effective'' correlation $\rho^{2}$ of these combined tracers with gravitational lensing:
\be
\rho^{2} = \sum_{i,j}\frac{C^{\kappa i}~(C^{-1})_{ij}~C^{\kappa j}}{C^{\kappa\kappa}}.
\label{eqn:rho-combined}
\ee

@@

\subsection{Improving efficiency with tomographic binning}\label{sec:tomo-bin}
Furthermore the delensing efficiency of galaxy surveys can be improved by taking into account redshift information.
When we weight a tracer with $\frac{C^{\kappa I}}{C^{II}}$ in \refeq{fweight} to maximize its ability to reconstruct the lensing potential we are only using redshift averaged information about the survey. However, as can be seen in \reffig{kernel}, the overlap varies as a function of redshifts. For this reason, the optimal approach is to weight galaxies at different redshifts with different weights according to both their cross-correlation with $\kappa$ and their auto-spectrum.
We can see this with a simple example. Let's split a single survey I into two non-overlapping redshift bins $I_{1}$ and ${I_{2}}$ with $I=I_{1}+I_{2}$.
For the full survey the effective cross correlation is equal to 
\be
\rho_{\rm{full}}^{2}= \frac{(\cl^{\kappa I})^2}{\cl^{\kappa \kappa} \cl^{I I}}=\frac{(\cl^{\kappa I_{1}}+\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} (\cl^{I_{1} I_{1}}+\cl^{I_{2} I_{2}})}
\ee
while for the split survey it will be 
\be
\rho_{\rm{split}}^{2}= \frac{(\cl^{\kappa I_{1}})^2}{\cl^{\kappa \kappa} \cl^{I_{1} I_{1}}}+\frac{(\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} \cl^{I_{2} I_{2}}}.
\ee
Now it can be show that $\rho_{\rm{split}}\ge \rho_{\rm{full}}$ being:
\be
\rho_{\rm{split}}^{2} - \rho_{\rm{full}}^{2}\propto  (\cl^{I_{1} I_{1}}\cl^{\kappa I_{2}}-\cl^{I_{2} I_{2}}\cl^{\kappa I_{1}})^2
\ee
Then $\rho^2$ is always bigger in the tomographic case, and the two are equal only when $\frac{C^{\kappa I_{i}}}{C^{I_{i}I_{i}}}$ is the same for all the redshift bins in which case a single optimal weight is sufficient for the entire survey.
Of course, this is an idealized situation: increasing the number of bins will reduce the number of galaxies in the bin thus increasing the shot noise. 
An optimal balance needs to be found given the survey specifics, but the general idea is that the optimal approach consists in binning the galaxies in the survey at different redshift. 
Of course, this is possible for galaxy surveys, but it is harder for a tracer with poor redshift information like the CIB or  radio continuum surveys.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel_tomo.pdf}
\caption{\textbf{Kernels Comparison with tomographic bins}: Comparison of the 10 LSST different kernels for every tomographic bin with the CMB lensing kernel. 
This will lead to a better delensing. }
\label{fig:kernel}

@@

In this section, we forecast the expected delensing efficiency and the relative importance of galaxy tracers for delensing in current and future experiments. We will use the Fisher information matrix to quantify the delensing efficiency as the improvement in the constraint of two inflationary parameters: the tensor to scalar ratio $r$ and the tensor tilt $n_{T}$.
We assume a CMB experimental scenario composed of a high-resolution CMB experiment which defines the internal reconstruction performance, together with a low-noise, low-resolution experiment the B-modes of which will be delensed and used to constrain the inflationary parameters.
We focus on three distinct scenarios: the current stage, a third generation stage (3G)  and finally the futuristic CMB Stage 4.

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/cmb_internal.pdf}
\caption{\textbf{CMB internal delensing:} Correlation factor between CMB internal reconstructed potential and the actual lensing potential for different CMB experiments and corresponding noise reconstructions. 
CMB measurements will rapidly improve providing the most efficient (even if biased) way to reconstruct the structures lensing the CMB.}
\end{center}
\label{fig:corr-internal}
\end{figure}

\subsection{Fisher method}
In the Fisher Information Matrix formalism \cite{10.2307/2342435}, the statistical uncertainty on a cosmological parameter $p$ can be obtain from the Fisher matrix $F_{ij}$ as $\sigma(p) =\sqrt{(\mathbf{F})_{pp}^{-1}}$.
We constrain the inflationary parameters $p =\{r,n_{t}\}$ with a CMB B-mode spectrum measurement so the Fisher matrix is:
\beq\label{eqn:fisher}
F_{ij} = \sum^{\ell_{max}}_{\ell=\ell_{min}}\frac{1}{\sigma( C_{\ell}^{BB} )^2}\frac{\partial {C}_{\ell}^{BB}}{\partial p_{i}}\frac{\partial {C}_{\ell}^{BB}}{\partial p_{j}}
\eeq
where we assume a Gaussian covariance:
\beq\label{eqn:cova}
\sigma( C_{l}^{BB} )= \sqrt{ \frac{2}{(2 {l}+1) \fsky }} \left( C_l^{BB,r} + C_{{l}}^{BB,\mathrm{lens}}+N_{{l}}^{BB} \right).
\eeq
The B-modes noise spectrum is given by \cite{knox:1995}:
\beq \label{eqn:noise}
N_l^{BB} = \left({\Delta_P}/{T_{\mathrm{CMB}}}\right)^2  e^{ {l^2 \theta_\mathrm{FWHM}^2}/({8 \ln 2})}
\eeq
where $\theta_\mathrm{FWHM}$ is the full half width of the telescope beam, and $\Delta_P$ is the instrumental noise of the experiment.
Even if the covariance structure has non-Gaussian contributions \cite{motloch:2017,benoit-levy:2012}, this approximation is good enough to show the improvement due to delensing.

It can be seen in \refeq{fisher} that removing the lensing contribution will improve parameters constraints.
The parameter uncertainties are inversely proportional to the covariance of the measurement. Since the lensing B-modes $C_{{l}}^{BB,\mathrm{lens}}$ are a substantial component of the covariance, removing part of them trough delensing will reduce the parameter statistical error.
After delensing the error can $\sigma^{\rm{del}}(p)$ can still be obtained from \refeq{fisher} with $C_{{l}}^{BB,\mathrm{res}}<C_{{l}}^{BB,\mathrm{lens}$ in the covariance defined in \refeq{cova}.
We define the improvement as the ratio of the constraints before and after delensing: $\alpha_{r} = \sigma^{\rm{del}}(r)/\sigma(r)$ and $\alpha_{n_{t}} = \sigma^{\rm{del}}(n_{t})/\sigma(n_{t})$.

In \refeq{fisher} we are making two important assumptions.
Firstly we are fixing the all the other cosmological parameters. Uncertainties in those will however propagate to uncertainties in $\{r,n_{t}\}$. While the constraints will be slightly optimistic, this has no real impact on the estimate of the improvement due to delensing which is the main focus of this work.
Furthermore we are not considering the uncertainties on galaxy surveys internal parameters such as biases, source distributions and photometric redshift uncertainties. These parameters do have a significant impact in degrading inflationary constraints \cite{sherwin:2015,namikawa:2016a}.
However as shown in \cite{sherwin:2015,namikawa:2016a} these can be auto-calibrated i.e. thy can be tightly constrained using galaxy surveys auto spectra and cross correlation.

@@

Very recently delensing has been proven possible on real data using as tracers both the internal CMB lensing potential reconstruction \cite{carron:2017} as well as CIB data \cite{larsen:2016a,manzotti:2017}.
Here we discuss the improvement due to combining these and other available surveys (see \cite{yu:2017} for publicly accessible multi-tracers data-products).

>>>>

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{natbib}
\newcommand{\bx}{{\bf x}}
\newcommand{\bk}{{\bf k}}
\newcommand{\bq}{{\bf q}}
\newcommand{\bP}{{\bf\Psi}}
\newcommand{\bs}{{\bf s}}
\newcommand{\cs}{{\cal S}}
\newcommand{\by}{{\bf y}}
\newcommand{\deltar}{\delta_{\rm recon}}%\def\lcdm{$\Lambda$CDM}
\def\arcsec{$^{\prime\prime}$}
\def\nl{N_\ell}
\def\sl{S_\ell}
\def\ublu{\bl'}
\newcommand{\conv}[2]{\left(\frac{#1}{#2}\right)}
\def\arctanh{\mathop{\rm arctanh}\nolimits}
%\renewcommand{\eqref}[1] {equation $($\ref{#1}$)$}
\newcommand{\comment}[1]{{}}
\def\beq{\begin{equation}}
\def\eeq{\end{equation}}
\def\beqn{\begin{eqnarray}}
\def\eeqn{\end{eqnarray}}
\def\a{\alpha}
\def\cl{C_{\ell}}
\def\h{\mathrm{h}}
\def\d{\rmn{d}}
\def\pa{\partial}
\def\deldel#1#2{\frac{\pa{#1}}{\pa{#2}}}
\def\ba{\bm{\alpha}}
\def\fracj#1#2{{\textstyle{#1\over#2}}}
\def\bxi{\bm{\xi}}
\def\half{\frac{1}{2}}
\def\ti{\widetilde}
\def\O{\Omega}
\def\OL{\Omega_\Lambda}
\def\Om{\ensuremath{\Omega_{\mathrm{m}}}}
\def\Ob{\ensuremath{\Omega_{\mathrm{b}}}}
\def\Oc{\ensuremath{\Omega_{\mathrm{CDM}}}}
\def\msol{\ensuremath{M_\odot}}
\def\l{\left}
\def\r{\right}
\def\o{\omega}
\def\gcm{\textrm{g cm$^{-3}$}}
\def\2gcm{\textrm{g cm$^{-2}$}}
\def\Scr{\Sigma_{\mathrm{crit}}}
\def\rcr{\rho_{\mathrm{crit}}}
\def\phidot{\ensuremath{\dot\phi_{\bl,\bl'}}}
\def\ddelta{\ensuremath{\dot\delta}}
\def\modu#1{\l |{#1}\r |}
\def\av#1{\l \langle{#1}\r \rangle}
\def\hmpc{\:{h}^{-1}\mathrm{Mpc}}
\def\th{\Theta}
\def\tth{\tilde\Theta}
\def\sg{\sigma}
\def\Sig{\Sigma}
\def\cf{{\cal F}}
\def\k{\kappa}
\def\P{{P}}
\def\pnl{{P}_{{\!\textrm{\tiny NL}}}}
\def\dnl{\Delta_{\text{\scriptsize NL}}}
\def\kmin{\k_{\mathrm{min}}}
\def\kmax{\k_{\mathrm{max}}}
\def\hires{the \emph{highRes} experiment}
\def\lores{the \emph{lowRes} experiment}
\def\ktot{\k_{\mathrm{tot}}}
\def\hunit{\ensuremath{\mathrm{km}{\mathrm{s}^{-1}} \mathrm{Mpc}^{-1}}}
\def\H0{\ensuremath{\mathrm{H}_0}}
\def\nn{\nonumber}
\def\lin{\mathrm{lin}}
\def\ISW{\mathrm{ISW}}
\def\bl{\bmm{l}}
\def\bL{\bmm{L}}
\def\fsky{f_{\mathrm{sky}}}
\newcommand{\E}[1]{\times 10^{#1}}
\newcommand{\bmm}[1]{{\mathbf{#1}}}
\newcommand{\bsection}[1]{\section{\uppercase{#1}}}

@@

For future experiments, we need to increase the delensing efficiency by almost a factor of 3 to fully exploit the expected instrumental capabilities \cite{abazajian:2016}.
In this paper, we propose and study a possible way: using future galaxy surveys as tracers of the lensing potential in addition to other probes such as the SKA radio continuum survey, cosmic infrared background and internal CMB reconstruction. 
Furthermore, we point out how using redshift information through tomographic binning can improve the delensing efficiency of galaxy survey.
These will translate into a better reconstruction of the B-modes in the measured patch, and, as a consequence will improve the constraints on inflation through delensing. We model several actual and future surveys, and after computing the residual B-modes, we forecast the resulting statistical uncertainties on the amplitude and the shape of the inflationary tensor perturbations for current CMB experiments as well as the next generation (S3) and the planned fourth generation (S4).

@@

The CMB lensing potential is \textit{the} field that we need to reconstruct to reverse the effect of large scale structure and delens the CMB.
However, the lensing potential can also be reconstructed using the CMB itself. In that case, we can treat it as a noisy tracer of the true field. Both the CMB lensing field and its noisy reconstructed counterpart have the same kernel $W^{\kappa}(z)$.
However, when computing the power spectrum of the latter, we need to add a noise component. Given the instrumental noise level and the beam, we can calculate the reconstruction noise $N_{\ell}^{\kappa \kappa}$, and so
\beq
\cl^{\kappa_\mathrm{rec} \kappa_\mathrm{rec}} = \cl^{\kappa \kappa } +N^{\kappa \kappa}_{\ell}
\eeq
In this work, the level of noise is computed assuming an iterative approach to the CMB lensing reconstruction as described in \cite{smith:2012,hirata:2003}.

@@

From this we get the power spectrum of the lensing component of the B-modes:
\be
C_{\ell}^{BB,\mathrm{lens}}  = \int \frac{d^2 \bl'}{(2 \pi)^2} W^2 (\bl,\bl')C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}.
\ee

The B-mode power spectrum measured on the sky is composed of a possible primordial component $C_{\ell}^{BB,r}$ together with the lensing $C_{\ell}^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_{\ell}^{BB}$ (defined in \refeq{noise}):
\be
C_{\ell}^{BB,\mathrm{full}} = C_{\ell}^{BB,r} + C_{\ell}^{BB,\mathrm{lens}} + N_{\ell}^{BB}.
\ee
The lensing component is a significant source of B-modes that, at large scales, correspond to a white noise source of roughly $5 \mu K$-arcmin. This means that it is not only bigger than the allowed inflationary component at scales smaller than several degrees, but it is also comparable with current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, we build a template \refeq{blens} of the lensing B-modes in the observed patch given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we can estimate $\phi$ using "tracers" of the dark matter distribution that sources the potential.

@@

The optimal weights $f(\bl,\bl')$ such that the residual lensing B mode power is minimized are:
\be
\label{eqn:fweight}
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }.
\ee
Here $C^{\kappa I}$ and $C^{II}$ are the cross-correlation spectrum of the tracer $I$ with the lensing convergence $\kappa$ and its autospectrum; they are described for each LSS field in \refsec{model}. The power spectrum of the E-modes noise $N^{EE}$ is the same as the B-modes one in \refeq{noise}.

@@

With this choice of $f(\bl,\bl')$ we finally have that the residual power is:
\beqn
\label{eqn:Bres}
C_{{l}}^{BB,\mathrm{res}} &=&   \int \frac{d^2 \bl'}{(2 \pi)^2}  W^2 (\bl,\bl')
 C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}  \\
&\times&  \left[1 - \left(\frac{C^{EE}_{{l'}} }{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right) \rho^2_{|\bl-\bl'|} \right] \nonumber
\eeqn
with
\be\label{eqn:rho1}
\rho_{\ell}= \frac{\cl^{\kappa I}}{\sqrt{\cl^{\kappa \kappa} \cl^{I I}}}.
\ee
The bigger the $\rho_{\ell}$ is for an LSS field, the more it is correlated with the lensing potential acting on the CMB photons. An higher correlation allows for a better reconstruction of $\phi$ and, as a consequence, of $B^{\rm{lens}}$.

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/BB_res.pdf}
\caption{}
\label{fig:bb_res}
\end{center}
\end{figure}

@@

We start by assuming that we have n different tracers of the gravitational potentials $I_{i}$ with $i\in \{1,..,n\}$. We can optimally combine them to estimate $\phi$ or, in other word, maximizing the correlation factor $\rho$ with:
\beqn
I         &=&    \sum_{i}c^{i}I^{i} \nonumber \\
c_{i}     &=& (C_{II}^{-1})_{ij}C^{\kappa I^{j}}
\label{eqn:combined}
\eeqn
where $C_{II}$ is the covariance matrix of the LSS tracers.
The residual B-mode power can be derived from \refeq{Bres} using an ``effective'' correlation $\rho^{2}$ of these combined tracers with gravitational lensing:
\be
\rho^{2}_{\ell} = \sum_{i,j}\frac{C^{\kappa i}_{\ell}~(C^{-1}_{\ell})_{ij}~C^{\kappa j}_{\ell}}{C^{\kappa\kappa}_{\ell}}.
\label{eqn:rho-combined}
\ee

@@

\subsection{Improving efficiency with tomographic binning}\label{sec:tomo-bin}
Furthermore the delensing efficiency of galaxy surveys can be improved by taking into account redshift information.
When we weight a tracer with $\frac{C^{\kappa I}}{C^{II}}$ in \refeq{fweight} to maximize its ability to reconstruct the lensing potential we are only using redshift averaged information about the survey. However, as can be seen in \reffig{kernel}, the overlap varies as a function of redshifts. For this reason, the optimal approach is to weight galaxies at different redshifts with different weights according to both their cross-correlation with $\kappa$ and their auto-spectrum.
We can see this with a simple example. Let's split a single survey I into two non-overlapping redshift bins $I_{1}$ and ${I_{2}}$ with $I=I_{1}+I_{2}$.
For the full survey the effective cross-correlation is equal to 
\be
\rho_{\rm{full}}^{2}= \frac{(\cl^{\kappa I})^2}{\cl^{\kappa \kappa} \cl^{I I}}=\frac{(\cl^{\kappa I_{1}}+\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} (\cl^{I_{1} I_{1}}+\cl^{I_{2} I_{2}})}
\ee
while for the split survey it will be 
\be
\rho_{\rm{split}}^{2}= \frac{(\cl^{\kappa I_{1}})^2}{\cl^{\kappa \kappa} \cl^{I_{1} I_{1}}}+\frac{(\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} \cl^{I_{2} I_{2}}}.
\ee
Now it can be show that $\rho_{\rm{split}}\ge \rho_{\rm{full}}$ being:
\be
\rho_{\rm{split}}^{2} - \rho_{\rm{full}}^{2}\propto  (\cl^{I_{1} I_{1}}\cl^{\kappa I_{2}}-\cl^{I_{2} I_{2}}\cl^{\kappa I_{1}})^2
\ee
Then $\rho^2$ is always bigger in the tomographic case, and the two are equal only when $\frac{C^{\kappa I_{i}}}{C^{I_{i}I_{i}}}$ is the same for all the redshift bins in which case a single optimal weight is sufficient for the entire survey.
Of course, this is an idealized situation: increasing the number of bins will reduce the number of galaxies in the bin thus increasing the shot noise. 
An optimal balance needs to be found given the survey specifics, but the general idea is that the optimal approach consists in binning the galaxies in the survey at different redshift. 
Of course, this is possible for galaxy surveys, but it is harder for a tracer with poor redshift information like the CIB or  radio continuum surveys.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel_tomo.pdf}
\caption{\textbf{Kernels Comparison with tomographic bins}: Comparison of the 10 LSST different kernels for every tomographic bin with the CMB lensing kernel. 
This will lead to a better delensing. }
\label{fig:kernel}
\end{center}
\end{figure}
% ===================

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/B_res_bin.pdf}
\caption{Tomographic bins improve the cross-correlation with lensing. Here we show the cross-correlation coefficient (\refeq{rho1}) as a function of scale for full surveys (solid lines) and tomographically binned surveys (dot-dashed line).}
\label{fig:bb_res}
\end{center}
\end{figure}

@@

In this section, we forecast the expected delensing efficiency and the relative importance of galaxy tracers for delensing in current and future experiments. We will use the Fisher information matrix to quantify the delensing efficiency as the improvement in the constraint of two inflationary parameters: the tensor to scalar ratio $r$ and the tensor tilt $n_{T}$.
We assume a CMB experimental scenario composed of a high-resolution CMB experiment which defines the internal reconstruction performance, together with a low-noise, low-resolution experiment the B-modes of which \textcolor{red}{(?)} will be delensed and used to constrain the inflationary parameters.
We focus on three distinct scenarios: the current stage, a third generation stage (3G)  and finally the futuristic CMB Stage 4.

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[scale=1.]{../images/cmb_internal.pdf}
%\caption{\textbf{CMB internal delensing:} Correlation factor between CMB internal reconstructed potential and the actual lensing potential for different CMB experiments and corresponding noise reconstructions. 
%CMB measurements will rapidly improve providing the most efficient (even if biased) way to reconstruct the structures lensing the CMB.}
%\end{center}
%\label{fig:corr-internal}
%\end{figure}

\subsection{Fisher Information Matrix}
In the Fisher Information Matrix formalism \cite{10.2307/2342435}, the statistical uncertainty on a cosmological parameter $p$ can be obtain the from inverse of the Fisher matrix $F_{ij}$ as $\sigma(p) =\sqrt{(\mathbf{F})_{pp}^{-1}}$.
We constrain the inflationary parameters $p =\{r,n_{t}\}$ with a CMB B-mode spectrum measurement so the Fisher matrix is:
\beq\label{eqn:fisher}
F_{ij} = \sum^{\ell_{max}}_{\ell=\ell_{min}}\frac{1}{\sigma( C_{\ell}^{BB} )^2}\frac{\partial {C}_{\ell}^{BB}}{\partial p_{i}}\frac{\partial {C}_{\ell}^{BB}}{\partial p_{j}}
\eeq
where we assume a Gaussian covariance:
\beq\label{eqn:cova}
\sigma( C_{\ell}^{BB} )= \sqrt{ \frac{2}{(2 {l}+1) \fsky }} \left( C_{\ell}^{BB,r} + C_{{l}}^{BB,\mathrm{lens}}+N_{{l}}^{BB} \right).
\eeq
The B-modes noise spectrum is given by \cite{knox:1995}:
\beq \label{eqn:noise}
N_{\ell}^{BB} = \left({\Delta_P}/{T_{\mathrm{CMB}}}\right)^2  e^{ {l^2 \theta_\mathrm{FWHM}^2}/({8 \ln 2})}
\eeq
where $\theta_\mathrm{FWHM}$ is the full half width of the telescope beam, and $\Delta_P$ is the instrumental noise of the experiment.

It can be seen in \refeq{fisher} that removing the lensing contribution will improve parameters constraints.
The parameter uncertainties are inversely proportional to the covariance of the measurement. Since the lensing B-modes $C_{{l}}^{BB,\mathrm{lens}}$ are a substantial component of the covariance, removing part of them trough delensing will reduce the parameter statistical error.
After delensing the error $\sigma^{\rm{del}}(p)$ can still be obtained from \refeq{fisher} with $C_{{l}}^{BB,\mathrm{res}}<C_{{l}}^{BB,\mathrm{lens}}$ in the covariance defined in \refeq{cova}.
We define the improvement as the ratio of the constraints before and after delensing: $\alpha_{r} = \sigma^{\rm{del}}(r)/\sigma(r)$ and $\alpha_{n_{t}} = \sigma^{\rm{del}}(n_{t})/\sigma(n_{t})$.

In \refeq{fisher} we are making a few important assumptions.
Firstly, we are fixing all the cosmological parameters except $\{r,n_{t}\}$. Uncertainties in those will, however, propagate to uncertainties in $\{r,n_{t}\}$. While neglecting this will lead to slightly optimistic constraints, it has no significant impact on the estimate of the improvement due to delensing.

We are also neglecting an important contribution to measured CMB B-modes: Galactic polarized foregrounds. 
The amplitude of these has been constrained in \cite{bicep2/keck-collaboration:2015, planck-collaboration:2015} and it strongly varies in different parts of the sky. Future experiments will use multi-frequency data to exploit the frequency dependency of these contaminants to remove them from the data. The amount of residual foregrounds depends both on uncertain foreground properties and experimental choices (see a review in \cite{abazajian:2016}).
For this reason, accurately treating foreground requires the use of simulations and knowledge about experimental details. 
We decided to focus on an ideal situation of no foregrounds or perfect cleaning even if the importance of delensing will be overestimated.

Furthermore, we are not considering the uncertainties on galaxy survey internal parameters such as biases, source distributions and photometric redshift uncertainties. These parameters do have a significant impact in degrading inflationary constraints \cite{sherwin:2015,namikawa:2016a}.
However, as shown in \cite{sherwin:2015,namikawa:2016a}, these can be auto-calibrated i.e. they can be tightly constrained using galaxy surveys auto spectra and cross-correlation. We checked this for a few of the tracers combinations used here and find it particularly true once several tracers are jointly taken into account.
Finally, we have assumed a Gaussian covariance even if the covariance structure has non-Gaussian contributions \cite{motloch:2017,benoit-levy:2012}. This approximation is good enough to show the improvement due to delensing.

@@

Only recently delensing has been performed on data using both the internal CMB lensing potential reconstruction \cite{carron:2017} as well as CIB maps \cite{larsen:2016a,manzotti:2017} as large scale structure tracers.
Here we discuss the improvement that can be obtained by combining these and other currently available tracers (see \cite{yu:2017} for publicly accessible multi-tracers data-products).
